Basic Model formulation

### 2015-01-30 documentation/model_formulation.Rmd
update model formulation. spatially-explicit model development

---
title: "Basic Model formulation"
author: "Florian D. Schneider"
date: "Tuesday, November 18, 2014"
output:
  html_document:
    keep_md: yes
  pdf_document:
    fig_height: 3.5
    fig_width: 4
fontsize: 12pt
geometry: margin=1.4in
csl: ../manuscript/amnat.csl
bibliography: ../manuscript/cas02.bib
---

@@

```{r, echo=FALSE}
G <- function(P, r = 2.25, K = 0.9) r*P*(1-P/K)

@@

$$ r = r_0 * b $$

@@

The

@@

@@

#### null model

@@

$$ C(V) = m*V + \frac{aVL}{1+ahV} \,,$$

@@

$$ \frac{\mathrm{d}\rho_1}{\mathrm{d}t} = r b V (1-\frac{V}{K}) - m V - \frac{aVL}{1+ahV} $$

@@

plot(ODEnullmodel, type = "l", ylim = c(0,1))

@@

The differential equations are formulated as changes in vegetation cover. Growth are positive changes, and deaths are negative changes.

Now, the CA uses transition probabilities. That is, a) the individual risk of death of vegetated cells and b) the individual chance of becoming vegetated for an empty cell.
To translate into the CA context, we need to find rules that apply to a) the vegetated cells, and b) the empty cells. Thus, it would be a) the reduction of term of the population of vegetated cells, divided by cover, to get the individual risk of death

@@

```{r, echo=FALSE}
par(mar = c(4,4,2,4))
plot(rho,  with(parms,  r * b * rho * ( 1 - (rho / K)) / (1 - rho) ) , ylab = "individual growth on empty cells", type = "l", ylim = c(0,2.5), las = 1, yaxs = "i")

@@

And the animated lattice would look like this:
```{r, eval = FALSE, echo = FALSE, warning = FALSE}
animateCA(nullmodel, "ca_nullmodel.gif")
```
![](ca_nullmodel.gif)

@@

These mechanisms can be implemented in the model by using a pair-approximation approach. 
In numerical simulations $G$ and $C$ can be implemented as the transition rates in a two-state spatial model.

In the formalism of the pair-approximation model of @kefi07, growth, *i.e.* the total amount of empty cells transitioning into vegetated cells, becomes a complex function of the relative cover of the multiple pairs including a $+$.

In a model with only global effects, the transition probability of indiviual plant death would be

$$ w_{1,0} = C(V)/V = m + \frac{aL}{1+ahV} $$

and the transition probability of individual plant growth

$$ w_{0,1} = G(V)/V = r (1-\frac{V}{K}) \,.$$

@@

This is assuming that the local environmental suitability is enhanced by a local facilitation term similar to the implementation in Kéfi et al 2007. Thus, the growth rate becomes a function of local vegetation cover $\nu$:

$$ r = r_0 + f \nu \,,$$

with an intrinsic growth rate $r_0$ which determines the probability of growth in absence of local vegetation and the additive facilitation $f$ which gradually determines the enhancement of the growth with an increasing local vegetation cover (maximizing at $\nu = 1$ if the cell has 4 neighbors).

@@

$$ K = K_0 (1 - c \nu)$$

@@

Mortality becomes enhanced in sparse environments by patches that are attracting grazers. We assume this attractivity, $v$, reduces search times locally.

$$ a = a_0 * (1 - v * \nu) $$

@@

Plants in grazed habitats develop protective traits, such as thorns or cushion growth. They thereby provide protected habitat to their direct neighborhood, or share the investments in those traits with their neighbors. Overall grazing mortality thus is reduced through an increase in handling time by associational resistance, $p$

$$ h = h_0 * p * \nu $$

### implementation in a cellular automaton

It is straigthforward to implement the model in a cellular automata model. It is defined on a defined grid of cells where each cell can take the values '1' if it is vegetated and '0' if it is empty. The transition rules between those states depend either on global vegetation cover $\rho_1$ or on the local cover in the direct neighborhood of the cell $\nu_1$ (closest 4 cells, von-Neumann-neighborhood).

Thus, two transition probabilities need to be defined: the probability of **growth**, $w_{0,1}$, turning an empty cell into a vegetated, and the probability of **death**, $w_{0,1}$, removing vegetation from the cell.

The null model, i.e. the model with only global effects on mortality and growth, with the basic assumption of a type II functional response and logistic growth, would thus be defined by

$$ w_{0,1} = r (1-\frac{\rho_1}{K})$$

@@

The pair-approximation framework (Matsuda et al 1992) is defined as a system of ordinary differential equations, describing the populations of *pairs* of cells. In a two state system, three pairs can be defined with population densities $\rho_{1,1}$ , $\rho_{0,0}$, and $\rho_{0,1}$. Besides, there are the population densities of singletons, $\rho_0$ and $\rho_1$.

@@

This means, the system can be fully described by two differential equations:

$$ \frac{\mathrm{d}\rho_{1}}{\mathrm{d}t} = $$

$$ \frac{\mathrm{d}\rho_0}{\mathrm{d}t} = $$

>>>>

---
title: "Basic Model formulation"
author: "Florian D. Schneider"
date: "Tuesday, November 18, 2014"
output:
  pdf_document:
    fig_height: 4
    fig_width: 6
  html_document:
    keep_md: yes
fontsize: 12pt
geometry: margin=1.4in
csl: ../manuscript/amnat.csl
bibliography: ../manuscript/cas02.bib
---

@@

@@

$$ r = r_0 b $$

@@

As an accelerating assumption on plant mortality we add the formulation of density dependent search efficiencies.

$$ b = b_0 N^q $$

Which turns the consumption term into

@@

Thus, plant mortality is accelerating with increasing cover, protecting plants at low cover, due to inefficient search behaviour, and then saturates at the inverse of handling time, $h$.

@@

#### consumption null model

@@

$$ C(V) = mV + \frac{aVL}{1+ahV} \,,$$

@@

$$ \frac{\mathrm{d}V}{\mathrm{d}t} = r b V (1-\frac{V}{K}) - m V - \frac{aVL}{1+ahV} $$

@@

plot(ODEnullmodel, type = "l", ylim = c(0,1))
axis(4, at = round(tail(ODEnullmodel,1)[["1"]],2),cex.axis = 0.8, las = 1 )

#### global model

The ODE that includes the additional global feed backs of water runoff and  reduced search efficience at low cover would read:

$$ \frac{\mathrm{d}V}{\mathrm{d}t} = r b V^{1 + \alpha} (1-\frac{V}{K}) - m V - \frac{aV^{1+q}L}{1+ahV^{1+q}} $$

We extend the parameter set by

parms <- list(
    m = 0.05,
    r = 1,
    b = 0.9, 
    K = 0.9, 
    a = 0.6,
    h = 100,
    L = 16, 
    alpha = 0.1,
    q = 1
  )

 as.data.frame(parms)
```

which gives us the prediction of complex bistability.

```{r, echo=FALSE}
plot(rho,  with(parms,  m * rho + (a * rho^( 1 + q) * L)/(1 + a * h * rho^( 1 + q))), ylab = "growth", type = "l", ylim = c(0,0.5))
  lines(rho, with(parms, r* b* rho^( 1 + alpha) * (1 - rho/K) ), col = "green3" )

d_rho <- function(rho, parms, z = 4) { 
  with(parms, r* b* rho^( 1 + alpha) * (1 - rho/K) - m * rho - (a * rho^( 1 + q) * L)/(1 + a * h * rho^( 1 + q)) )  
  }

# running the ode-solver
ODEglobal_high <- ode(y = 0.9, func = odesys, times = 1.1^(1:50), parms = parms)

# transfer into ouput and calculate missing rho values
ODEglobal_high <- as.data.frame(ODEglobal_high)

plot(ODEglobal_high, type = "l", ylim = c(0,1))
axis(4, at = round(tail(ODEglobal_high,1)[["1"]],2),cex.axis = 0.75, las = 1 )

# running the ode-solver
ODEglobal_low <- ode(y = 0.1, func = odesys, times = 1.1^(1:50), parms = parms)

# transfer into ouput and calculate missing rho values
ODEglobal_low <- as.data.frame(ODEglobal_low)
axis(4, at = round(tail(ODEglobal_low,1)[["1"]],2),cex.axis = 0.75, las = 1 )

lines(ODEglobal_low)

@@

It is straigthforward to implement the model in a cellular automata model. It is defined on a defined grid of cells where each cell can take the values '1' if it is vegetated and '0' if it is empty. The transition rules between those states depend on global vegetation cover, $\rho_1$.

While the differential equations are formulated as changes in vegetation cover(growth are positive changes, and deaths are negative changes), the CA uses transition probabilities between the two possible cell states. Thus, two transition probabilities need to be defined: the probability of **growth**, $w_{0,1}$, turning an empty cell into a vegetated, and the probability of **death**, $w_{0,1}$, removing vegetation from the cell.

To translate into the CA context, we need to find rules that apply to a) the vegetated cells, and b) the empty cells. Thus, it would be a) the reduction of term of the population of vegetated cells, divided by cover, to get the individual risk of death

@@

parms <- list(
    m = 0.05,
    r = 1,
    b = 0.9, 
    K = 0.9, 
    a = 0.6,
    h = 100,
    L = 10, 
    alpha = 0,
    q = 0
  )

par(mar = c(4,4,2,4))
plot(rho,  with(parms,  r * b * rho * ( 1 - (rho / K)) / (1 - rho) ) , ylab = "individual growth on empty cells", type = "l", ylim = c(0,2.5), las = 1, yaxs = "i")

@@

And the animated lattice would look like this:
```{r, eval = FALSE, echo = FALSE, warning = FALSE}
animateCA(nullmodel, "ca_nullmodel.gif")

![](ca_nullmodel.gif)

#### additional feedbacks

The next step is to include the additional global feedback terms into the model. This is straightforward

$$ w_{1,0} = m + \frac{aL\rho_1^{q}}{1+ah\rho_1^{1+q}} $$

$$ w_{0,1} = \frac{r \rho_1^{1+\alpha} b (1-\frac{\rho_1}{K})}{1-\rho_1} $$

parms <- list(
    m = 0.05,
    r = 1,
    b = 0.9, 
    K = 0.9, 
    a = 0.6,
    h = 100,
    L = 16, 
    alpha = 0.1,
    q = 1
  )

set.seed(5535732) # setting seed for random number generation

globalmodel <- runCA(0.9, parms, delta = 0.2, t_max = 150, t_min = 100, t_eval = 50, isstable = 0.0001, saveeach = 1)

plot(globalmodel$rho_one ~ globalmodel$time, 
     ylim = c(0,1), ylab = expression( rho["+"]), 
     xlab = "time [yr]", type = "l")

globalmodel <- runCA(0.1, parms, delta = 0.2, t_max = 150, t_min = 100, t_eval = 50, isstable = 0.0001, saveeach = 1)
lines(globalmodel$rho_one ~ globalmodel$time)
lines(ODEglobal_low, lty = 3)
lines(ODEglobal_high, lty = 3)
```

\newpage

@@

These mechanisms can be implemented in the model by using a pair-approximation approach and numerical simulations, where $G$ and $C$ can be implemented as the transition rates in a two-state spatial model depending on global and local vegetation cover.

@@

This is assuming that the local environmental suitability is enhanced by a local facilitation term similar to the implementation in Kéfi et al 2007. Thus, the reduction in growth rate, $r$, by aridity, $a$, is compensated by a function of $q_{0|1}$, the local density of cells in state 1 (i.e. vegetation) given that the focal cell is in state 0 (i.e. empty), and maximizes to one if the cell has four neighbors:

$$ a = a_0 + (f - a_0) q_{0|1} \,,$$

with the effect of aridity, $a_0$ in absence of local vegetation being increased to the value of $f$ if additive facilitation due to the presence of neighboring plants occurs. This term gradually determines the enhancement of the growth with an increasing local vegetation cover (maximizing at $q_{0|1} = 1$ if the cell has 4 neighbors).

@@

$$ K = K_0 (1 - c q_{0|1})$$

That is, if a cell has a fully vegetated local neighborhood, it's carrying capacity will be reduced by $c$. Thus, $c \geq K$ otherwise carrying capacity will fall below $0$ at high cover.

@@

Mortality becomes enhanced in sparse environments by patches that are attracting grazers. We assume this attractivity enhances search efficiency locally by the value $v$

$$ b = b_0  + v q_{1|1} $$

Thus, at low local cover feeding will be high on plants with neighbors, whereas at high cover the term has no effect since handling time is limiting consumption. An enhanced search efficiency will raise the critical thresholds for a collapse.

@@

Plants in grazed habitats develop protective traits, such as thorns or cushion growth. They thereby provide protected habitat to their direct neighborhood, or share the investments in those traits with their neighbors. Overall grazing mortality thus is reduced through an increase in handling time by associational resistance, $p$,

$$ h = h_0  / (1 - p  q_{1|1}) $$

An alternative formulation should be found. The term is highly non-linear and provokes relevant changes only at values above 0.9. It is defined within $(0,1($

@@

The pair-approximation framework [@matsuda92] is defined as a system of ordinary differential equations, describing the populations of *pairs* of cells. In a two state system, three pairs can be defined with population densities $\rho_{1,1}$ , $\rho_{0,0}$, and $\rho_{0,1}$. Besides, there are the population densities of singletons, $\rho_0$ and $\rho_1$.

@@

Since formally $q_{i|j}$ is a conditional probability (given that one neighbor is in state $i$, what is the probability to find the other in state $j$) we can express it in terms of densities of pairs and ‘singletons,’ as

$$ q_{i|j} = \frac{\rho_{ij}}{\rho_i} $$

The vegetation dynamics of the system can be fully described by two differential equations (omitting indexed zeros of equations for easier reading):

$$ \frac{\mathrm{d}\rho_1}{\mathrm{d}t} = r ( a + (f - a) q_{0|1} ) \rho_1^{1 + \alpha} (1-\frac{\rho_1}{K (1 - c q_{0|1})}) - m \rho_1 - \frac{(b  + v q_{1|1} )\rho_1^{1+q}L}{1+( b  + v q_{1|1} ) ( h  / (p  q_{1|1}))\rho_1^{1+q}} $$

$$ \frac{\mathrm{d}\rho_0}{\mathrm{d}t} =  m \rho_1 + \frac{(b  + v q_{1|1} )\rho_1^{1+q}L}{1+( b  + v q_{1|1} ) ( h  / (p  q_{1|1}))\rho_1^{1+q}} - r ( a + (f - a) q_{0|1} ) \rho_1^{1 + \alpha} (1-\frac{\rho_1}{K (1 - c q_{0|1})})$$

Next steps:

- formulate equations for $\frac{\mathrm{d}\rho_{0,1}}{\mathrm{d}t}$, $\frac{\mathrm{d}\rho_{0,0}}{\mathrm{d}t}$, and $\frac{\mathrm{d}\rho_{1,1}}{\mathrm{d}t}$

### implementation in CA

It is quite straightforward to implement the mechanisms described above in the syntax of a cellular automata model, where the probabilities of transition can not only depend on the global cover $\rho_1$ but also on the local cover in the direct neighborhood of the cell $\nu_1$ (closest 4 cells, von-Neumann-neighborhood).

see implementation in NetLogo: `documentation/livestockmodel.nlogo`
